---
title: "Trees and class intersections"
subtitle: QUESTIONS
output:
  pdf_document: default
  html_document:
    df_print: paged
geometry: margin=.75in
graphics: yes
header-includes:
- \usepackage{graphicx}
- \usepackage{color}
fontsize: 9pt
classoption: letter
---

```{r, setup, echo=FALSE}
library(knitr)
# Any of these options can be changed in any R chunk.
knitr::opts_chunk$set(echo = TRUE, 
                      warning = FALSE,
                      message = FALSE,
                      fig.align = "center", 
                      # Two following determine width and height
                      # of the R device on which the plots are made
                      fig.width = 7, 
                      fig.height = 6,
                      # last argument here determines the actual 
                      # height of the plot as it appears in the processed
                      # RMarkdown file
                      out.height = "50%") 

dataDirectory <- "./data"
imgDirectory <- "./img"
```

```{r}
library(tidyverse)
```

**28 marks**

In the previous question "Trees and pipelines" you constructed two data sets, `trees_DV` and `trees_EDA`, containing the results from students in the courses 
"Data Visualization" and "Exploratory Data Analysis". These were the results from the quiz described in the file "TreesContext".

Some students were in both courses.  To explore this a little more, analogous data sets to `trees_DV` and `trees_EDA` were constructed but with the (fake) student identification numbers attached.  

This data is contained in the files
"trees_DV_id.csv" and "trees_EDA_id.csv".  
Read these data in and assign the values to `trees_DV-id` and `trees_EDA_id` respectively.

In this question, the results of students in both course (who therefore took the quiz twice) will be explored.

In all parts, show your code unless told otherwise.

a. **(2 marks)**  Read the data in (with `read_csv()`) and assign the values to `trees_DV-id` and `trees_EDA_id` respectively. Show the first two rows of each data set.

   ```{r, results = 'hide'}
(trees_DV_id <- read_csv(file.path(dataDirectory, "trees_DV_id.csv")))
(trees_EDA_id <- read_csv(file.path(dataDirectory, "trees_EDA_id.csv")))
   ```
   ```{r}
head(trees_DV_id, 2)
head(trees_EDA_id, 2)
   ```

\newpage

b. **(9 marks)**  A focus on those students in both courses.

   i. *(2 marks)* Determine which students took the quiz twice.

      Show the number of such students and their ids.

   ```{r}
intersect(trees_DV_id$id, trees_EDA_id$id)
   ```
      
      \newpage
   
   i. *(2 marks)* Use an appropriate `"join"` function to construct a single data set called `trees_2` which contains **only** those students who answered the quiz for **both** courses.
   
      `trees_2` should contain `id` as well as all variables associated with each course (a total of 9 variables).
   
      Assign the result to `trees_2` and print the result.

   ```{r}
trees_2 <- inner_join(trees_DV_id, trees_EDA_id, by = "id")
trees_2
   ```
      \newpage
      
   i. *(5 marks)* Determine whether these students were asked the same question or not and whether they gave the same answers or not.

      Effect this by using a pipeline beginning with `trees_2`.
      
      It should return a `tibble` 
       
      - having only three variates: `id`, `same_question`, and `same_answer`
        
        - `same_question` is a logical vector indicating whether that 
          student was presented with the same question in both courses
          or not
          
        - `same_answer` is a logical indicating for those who had the 
          same question whether they provided identical answers or not.
          For those who had different questions, this should be missing.
      
      - having rows ordered by ascending `id`
      
      Assign the result to `trees_common` and print the result.
   
   ```{r}
# id, same_question, same_answer
trees_2 %>%
   transmute(id = id,
             same_question = (if_else(anchor.x == anchor.y, TRUE, FALSE)),
             same_answer = (if_else(same_question == FALSE, NA, if_else(height.x == height.y, TRUE, FALSE)))) %>%
   arrange(id) ->
   trees_common
trees_common
   ```
\newpage

c. **(8 marks)**  Putting all students together.
   
   
   i. *(2 marks)* Combine `trees_DV_id` and `trees_EDA_id` so that 
      no new columns are added, none are deleted, and the answers 
      from the `class` `"DataViz"` appear above those from the class `"EDA"`.
      
      Save the combined data as `trees_all` and evaluate the
      following:
      
      ```{r, eval = FALSE}
      dim(trees_all)
      head(trees_all, 2)
      tail(trees_all, 2)
      ```
   ```{r}
trees_all <- full_join(trees_DV_id, trees_EDA_id)
dim(trees_all)
head(trees_all, 2)
tail(trees_all, 2)
   ```
      \newpage
   
      
   
   i. *(6 marks)* Using only `trees_all` and `trees_common`,
      construct a data set  called `trees_work` such that it 
      
      - has only variables `id`, `class`, `anchor`, `greater`, and `height` (in that order)
      - has rows sorted in ascending order of `id` number, then `class`
      - contains no data that was duplicated (i.e. same student,  question, and answers)
      - if a student answered twice, with the same questions and the same answer, only one answer
        is used and the `class` is changed to `"BOTH"`
      - contains each set of answers from a student, provided they are different answers if to the same questions
      
      The data set `trees_work` **must** be constructed in a **single** pipeline with **no** intermediate assignment
      
      When done, evaluate the following:
      
      ```{r, eval = FALSE}
      # To test your answer the following should return TRUE
      (nrow(trees_work) + 
         sum(trees_common$same_answer, na.rm = TRUE)  == nrow(trees_all))
      #
      dim(trees_work)
      head(trees_work, 2)
      tail(trees_work, 2)
      ```
   ```{r}
full_join(trees_all, trees_common, by = "id") %>%
   arrange(id, class) %>%
   mutate(class = if_else(!is.na(same_question) | !is.na(same_answer), if_else(same_question==TRUE & same_answer==TRUE,"BOTH",class), class)) %>%
   unique() %>%
   select(-same_question, -same_answer) ->
   trees_work
(nrow(trees_work) + 
         sum(trees_common$same_answer, na.rm = TRUE)  == nrow(trees_all))
dim(trees_work)
head(trees_work, 2)
tail(trees_work, 2)
   ```

   
      
   \newpage

d. **(9 marks)** Analysis of the `trees_work` data just constructed.

   i. *(2 marks)* Are any of the answers provided by students contradictory?
      If so, which student(s) and what is contradictory.
      
      If not, why not?
     
      Show code and results to support your answer.
      
- Yes, there are answers that are provided by students contradictory.
      
   ```{r}
trees_work %>%
   filter(anchor > height & greater == TRUE | anchor < height & greater == FALSE)
   ```
     
      \newpage
      
   i. *(7 marks)* Using the `trees_work` data  with `ggplot2`, draw a **single** histogram of the estimated heights.  The plot should have
      
      - ggplot be fed the data via a pipe
      - have a title
      - have fill colour "white" with "black" border
      - have bins = 35
      - have a 50% transparent "grey" filled density estimate overlaid
      - have a vertical "black" line dashed line at each anchor point
      - a vertical "red" solid line at the actual height 
        of the tallest tree  ("Hyperion" at 115.7 metres)
      - the plots facetted by anchor and stacked vertically
        
   ```{r}
trees_work %>%
   ggplot(data = ., mapping = aes(x = height)) + aes(y = ..density..) + 
   geom_histogram(bins = 35, fill = "white", col = "black") +
   geom_density(fill = "grey", alpha = 0.5) +
   geom_vline(xintercept = 50, col = "black", linetype="dashed") +
   geom_vline(xintercept = 150, col = "black", linetype="dashed") + 
   geom_vline(xintercept = 115.7, col = "red") + 
   facet_wrap(~ anchor, ncol = 1) + 
   ggtitle("DataViz and EDA Tree data")
   ```

     