---
title: "Trees and pipelines"
subtitle: QUESTIONS
output:
  pdf_document: default
  html_document:
    df_print: paged
geometry: margin=.75in
graphics: yes
header-includes:
- \usepackage{graphicx}
- \usepackage{color}
fontsize: 9pt
classoption: letter
---

```{r, setup, echo=FALSE}
library(knitr)
# Any of these options can be changed in any R chunk.
knitr::opts_chunk$set(echo = TRUE, 
                      warning = FALSE,
                      message = FALSE,
                      fig.align = "center", 
                      # Two following determine width and height
                      # of the R device on which the plots are made
                      fig.width = 7, 
                      fig.height = 6,
                      # last argument here determines the actual 
                      # width of the plot as it appears in the processed
                      # RMarkdown file
                      out.width = "50%") 

dataDirectory <- "./data"
imgDirectory <- "./img"
```

```{r setup libraries, echo = FALSE}
library(tidyverse)
library(magrittr)
library(dplyr)
library(ggplot2)
```

**40 marks**

Recall, that in an earlier assignment a number of questions were asked about student guesses of the tallest coast redwood tree.
The entire context can be found in the again in the `TreesContext` file.

The results of the online quiz were summarized in the file `trees.csv` for analysis, but this is not how the data were provided by the online quiz service.  
Instead, they provided as two different ".csv" files, one for each course, in quite different structure and containing lots of extraneous information.

Anonymized versions of these data are given in the files:

- `DataViz_trees_anon.csv` containing the results from the students in `Data Visualization`, and
- `EDA_trees_anon.csv` containing the results from the students in `Exploratory Data Analysis`

The purpose of the present question is to exactly reproduce the data of `trees.csv` by reading in and reshaping the data in the above two files.  But first, there will be a little exploration of the data as it comes in.

- **All programming** to accomplish this reshaping is to be use `readr` functions for inputting data and `magrittr` **pipelines** to reshape the data.

- In answering all questions, **show your code** as part of your answer **unless** you are specifically asked not to.

- Whenever asked to print the result, just use the default printing (which will only print a subset of the result).  We don't need to see the whole thing, just whatever `R` prints.

a. **(2 marks)**  Using `read_csv()`, read in the data 

   - from `EDA_trees_anon.csv`, assigning the result to `EDA_raw`, and
   - from `DataViz_trees_anon.csv` assigning the result to  `DV_raw`.
   
   ```{r, results = "hide"}
(EDA_raw <- read_csv(file.path(dataDirectory, "EDA_trees_anon.csv")))
(DV_raw <- read_csv(file.path(dataDirectory, "DataViz_trees_anon.csv")))
   ```

\newpage   

b. **(12 marks)**  Some practice with pipes.

   i. *(3 marks)* The variable `"Org Defined ID"` contains the (anonymized) student id number.
      
      Construct a pipeline beginning with `EDA_raw`. It should
       
      - change the name of the variable `"Org Defined ID"` to simply `ID`,
      - determine the number of students answering the quiz, and
      - print the result
      
   ```{r}
EDA_raw %>%
   rename(ID = `Org Defined ID`) %>%
   select(ID) %>%
   n_distinct()
   ```
   i. *(3 marks)*  Construct a pipeline beginning with `EDA_raw`. It should
       
      - determine the number of students who started answering the quiz in the `AM`,
      - as well as the number in the `PM`, and then
      - print the result showing both numbers
   ```{r}
EDA_raw %>%
   distinct(`Org Defined ID`, .keep_all = TRUE) %>%
   filter(grepl("AM", `Attempt Start`)) %>%
   select(`Attempt #`) %>%
   summarise(countAM = n(), countPM = 36-countAM)
   ```
      
   i. *(3 marks)* The function `str_extract(x, "[0-9]+")` will extract the first set of contiguous digits from x.
   
      Construct a pipeline beginning with `EDA_raw`. It should
       
      - change the name of `"Answer Match"` to `answer`,
      - find all the distinct answers, 
      - change these answers to be the first set of contiguous digits, and
      - print the result
      
   ```{r}
EDA_raw %>%
   rename(answer = `Answer Match`) %>%
   distinct(answer) %>%
   str_extract("[0-9]+")
   ```
      
   i. *(3 marks)* The function `substring(x, first, last)` will extract from `x` the contiguous characters from the position identified by `first` to that defined by `last` (default value of `last` is `1000000L`).  Another function `toupper(x)` will convert the characters in `x` to be upper case (only affects letters).
   
      Construct a pipeline beginning with `DV_raw` (from the Data Visualization class). 
      
      It should
       
      - change the name of `"Answer Match"` to `answer`,
      - find all the distinct answers, 
      - select the first 5 characters of `answer`
      - apply `toupper` to `answer`
      - print the result
    
   ```{r}
DV_raw %>%
   rename(answer = `Answer Match`) %>%
   distinct(answer) %>%
   mutate(answer = substring(answer, first = 1, last = 5)) %>%
   toupper()
   #mutate(answer = toupper(substring(answer, first = 1, last = 5))) -> this returns col of dataframe
   ```

\newpage   

c. **(26 marks)**  Now you will construct pipelines to produce the data sets `trees_EDA`, and `trees_DV`.  These will then be combined to produce `trees_both` containing results from both classes.  In the end, your data set `trees_both` should exactly match the `trees` data set from a previous assignment.

   Each of these data sets will have exactly the followning four variables (in this order):
   
   - `class` -- a character vector identifying the class as either `"DataViz"` or `"EDA"`
   - `anchor` -- a numeric `double` vector containing the value of the anchor used in the question
   - `greater` -- a logical vector indicating whether the student answered that the height was greater than (`TRUE`) or less than (`FALSE`) the value given by the anchor
   - `height` -- the value guessed by the student

   The following information may be helpful in constructing your pipeline
   
   - elements of `"Q Title"` identify which context (either `"Trees 1"` or `"Trees 2"`),
     that is set of questions, is being asked.
   - the text of the questions (including the anchor value) is contained in `"Q Text"`
   - which question (first or second) is being answered in `"Answer"`
   - the answers are contained in `"Answer Match"`
   - if, when answering the first question, anyone writing `"less than"` should have answered `"A"`, and anyone writing  `"more than"`, or `"greater than"`, should have answered `"B"`.
   - if, when answering the first question, they wrote the value of the anchor, then that should be interpreted as answering `"A"` 
   - for the height guessed, the first numerical answer they give is to be used (e.g. take just the first value of a range)
  
   
   i. *(20 marks)* Beginning with `EDA_raw`, construct a single pipeline to produce `trees_EDA`.
    
      Show the result `trees_EDA`

   ```{r}
EDA_raw %>%
   filter(Difficulty == 1) %>%
   mutate(anchor = factor(if_else(`Q Title` == "Trees 1", 50, 150))) %>%
   pivot_wider(names_from = Answer, values_from = 'Answer Match') %>%
   transmute(class = "EDA", 
             anchor = anchor,
             greater = factor(if_else(`Answer for blank # 1` == "A", "FALSE", "TRUE")),
             height = (if_else(`Answer for blank # 2` == "100-110m", 100,extract_numeric(`Answer for blank # 2`)))
   ) %>%
   mutate(height = factor(if_else(`height` == 115.92, 115, height))) ->
   trees_EDA
trees_EDA
   ```

   i. *(4 marks)* Reuse your code to construct `trees_DV` from `DV_raw`
    
      - identify where your previous code is changed with a comment like:
      
        `# <----  change`
        
        at the end of the line where the change occurred.
        
      Show the result `trees_DV`

   ```{r}
DV_raw %>%
   filter(Difficulty == 1) %>%
   mutate(anchor = factor(if_else(`Q Title` == "Trees 1", 50, 150))) %>%
   pivot_wider(names_from = Answer, values_from = 'Answer Match') %>%
   mutate(blank1 = toupper(substr(`Answer for blank # 1`, 1, 1))) %>%            # <----  change
   transmute(class = "DataViz", 
             anchor = anchor,
             greater = factor(if_else(blank1 == "B", "TRUE", "FALSE")),      # <----  change
             height = extract_numeric(`Answer for blank # 2`)           # <----  change
   ) ->
   trees_DV
trees_DV
   ```
   

   i. *(2 marks)* Combine the two data sets by appending to the bottom of 
      `trees_DV` the rows of `trees_EDA` to create `trees_both`.
      
      Read in the values of `trees.csv` (with `read_csv()`) and 
      assign the result to `trees`.
      
      Check that these are identical by evaluating
      ```{r, eval = FALSE}
      sum(trees_both != trees)
      ```
      This should be 0.
      
   ```{r, results = "hide"}
trees_both <- rbind(trees_DV, trees_EDA)
trees <- read_csv(file.path(dataDirectory, "trees.csv"))
   ```
   ```{r}
sum(trees_both != trees)
   ```
      